Ensemble Stability of Biological Networks

### 2014-09-16 paper.tex
simplify characterization of stability

In order to accommodate the possibility of uncertainty in our modelling,
we will generalize our notion of a system on a network to that of a
system with random parameters.  Again, this will involve three
steps. First, we provide a measure space $S, \mathcal{B}(S)$ which represents the
values over which our parameters can vary. Next, instead of a single
vector field, we consider a family of vector fields parameterized by
this space. That is to say, we introduce a mapping $V \colon S \to \Gamma T(\prod_i M_m)$.
(Here $\Gamma$ denoted the space of sections of the tangent bundle.)
As before, for each point $p \in S$, the vector field $V(p)$ must be consistent with
the system graph. Finally, we select a probability measure $\mu \colon
\mathcal{B}(S) \to [0,1]$ which represents our understanding of which values of the parameters
are likely. In accord with Bayesian statistics, we might revise this
distribution as data comes in or use it to estimate parameters and error
bars from experimental results.

Having done this, we are now in a position to do a probabilistic analysis.
Given some quantity $q$ characterizing our system, this quantity now
becomes a random variable.  This variable may be discrete of continuous
depending on the quantity under consideration.  For example, if the quantity
is the time it takes for a particle to travel between two points, we have
a real-valued random variable; if the quantity is the number of fixed
points, we have an integer-valued random variable; and if the quantity is
whether or not the system posesses a limit cycle, we have a binary random
variable.

@@

\begin{lemma}
Given a polynomial $p(x) = z^n + \sum_{k=1}^{n-1} s_k x^{n-k}$ and a real number
$\epsilon > 0$, there exists $\delta > 0 \in \mathbb{R}$ such that, if $q(x) =
x^n + \sum_{k=1}^{n-1} \sigma_i x^i$ is a polynomial for which $|s_i - \sigma_i|
< \delta$ then one can order the roots of $p$ and $q$ such that $p(z) = \prod_{k=1}^n
(z - r_i)$ and $q(z) = \prod_{k=1}^n (z - \rho_i)$ and $| r_k - \rho_k | < \epsilon$.
\end{lemma}

\begin{proof}
Without loss of generality, we may assume that $\epsilon < 1$.

We first consider the case where $p(x) = x^n$.

If we make the rescaling $x = \epsilon y/2$ in the equation $p(x)  = 0$,
we obtain the equation
\[
y^n + \sum_{k=1}^{n-1} \left(\frac{2}{\epsilon}\right)^k s_k y^{n-k} = 0 .
\]
Cauchy's bound on the roots of this equation is
\[
|y| < 1 + \max_j \left(\frac{2}{\epsilon}\right)^j |s_j| ,
\]
which becomes the bound
\[
|x| < \frac{\epsilon}{2} + \max_j \left(\frac{2}{\epsilon}\right)^{j-1} |s_j|
< \frac{\epsilon}{2} + \left(\frac{2}{\epsilon}\right)^{n-1} \max_j |s_j|
\]
for the original equation.  Hence, if we choose $\delta = (\epsilon/2)^{n}$,
we will have $|x| < \epsilon$.

Next, we generalize to the case $p = (x-r)^n$.

To extend this result of the general case, we note that, given two
polynomials $p = x^m + \sum_{k=1}^n a_k x^{m-k}$ and
$q = x^n + \sum_{k=1}^n b_k x^{n-k}$, the Jacobian of the map which
sends the coefficients of the matrices to the coefficients of the
product $pq$ is the Sylvester matrix, whose determinant is the resultant
of $p$ and $q$.  Hence, is $p$ and $q$ have no common root, the map
is locally invertible.
\end{proof}

@@

\begin{proof}
Suppose that none of the real parts of the roots of a polynomial $p$ equal
zero.  Let $\epsilon$ be the minimum of the absolute values of the real
parts of the roots of $p$.  Then, by continuity of the map which sends the
roots of a polynomial to its coefficients, there exists a $\delta > 0$
such that, if $q$ is any polynomial whose coefficients differ by less than
$\delta$ from the

Suppose that we have a family
of polynomials whose roots are of one type but turn into another type.
There are three ways this could happen:
\begin{itemize}
\item{1.} A real root could go from being positive to being negative
  or vice versa.  For this to happen, we must pass through an
  intermediate value with a zero root.  For that intermediate value,
  we have $D = 0$.
\item{2.} A pair of real roots could turn into a pair of complex roots
  or vice versa.  For this to happen, we must pass through an
  intermediate value with a repeated root.  For that intermediate
  value, we have $\delta = 0$.
\item{3.} A conjugate pair of complex roots could go from having
  positive real part to having negative real part or vice versa.  For
  this to happen, we must pass through an intermediate value in which
  the real part is zero.  For that intermediate value, the two roots
  are imaginary so, bing conjugates, they sum to zero, hence we have
  $\Delta = 0$.
\end{itemize}
\noindent  Thus, we conclude that the boundaries of our regions
$R_{abcd}$ consist of portions of the surfaces $D=0, \delta=0,
\Delta=0$ and we will characterize them by studying these surfaces.

To complete the proof, we show that the condition $\delta = 0$ is not
needed to describe the boundary.  Suppose that $\delta = 0$.  Then there
must exist a repeated root.  If the real part of any root, repeated or
not, is positive, then the polynomial is not stable and, by continuity,
if one allows the coefficients to vary within a sufficiently small
ball about that solution, the root will remain negative, so the
polynomial cannot lie on the boundary of the stable region.  If all
the roots, including the repeated ones, are negative then, by a similar
argument the point is in the interior of the stable region.  The only
other possibility is that some root has zero real part, but then
$\Delta = 0$ so the condition $\delta = 0$ is superfluous.
\end{proof}

Introduce $V$ as the parameter space of all polynomials of degree $n$
with real coefficients.  In other words, $V$ is a real vector space of
dimension $n$ whose coordinates we will identify with the coefficients
$s_k$ of our polynomial.  We now partition this space into regions
corresponding to specific types of polynomials.  Let $R_{abcd}$ be the
subset of $V$ such that $x \in R_{abcd}$ if and only if the polynomial
corresponding to $x$ has $a$ positive real roots, $b$ negative real
roots, $c$ complex roots with positive real part and $d$ complex roots
with negative real part.  Once we characterize these regions using
invariants, we will have solved the original problem because a matrix
is stable if the point of $V$ corresponding to it lies in a region of
the form $R_{a0c0}$.

Next, we introduce some products:
\begin{align}
D &= \prod_{k=1}^n r_n \cr
\delta &= \prod_{j=1}^{n-1} \prod_{k=j+1}^n (r_1 - r_j)^2 \cr
\Delta &= \prod_{j=1}^{n-1} \prod_{k=j+1}^n (r_1 + r_j) \cr
\end{align}

Since these quantities are invariant under permutation of the roots
$r_j$, they may be expressed as functions of the coefficients $s_j$.
The explicit expressions may be obtained by such techniques as long
division and determinants. These quantities are of interest because they describe the boundaries between the regions introduced above.

The simplest case is $n=2$.  There we have the regions $R_{2000},
R_{1100}, R_{0200}, R_{0020}, R_{0002}$ and our invariants are $D =
s_2, \delta = s_1^2 - 4s_2, \Delta = s_1$.  The relation between these
is summarized by the conditions
\begin{align}
R_{2000} \qquad &D > 0, \delta > 0, \Delta > 0, \cr
R_{1100} \qquad &D < 0, \delta > 0, \cr
R_{0200} \qquad &D > 0, \delta > 0, \Delta < 0, \cr
R_{0020} \qquad &\delta < 0, \Delta > 0, \cr
R_{0002} \qquad &\delta < 0, \Delta < 0. \cr
\end{align}
These regions used for the classification of stability of two-component systems are show in \ref{fig:region2x2}. The stable regions are given by $R_{2000}$ and $R_{0020}$. All of the other regions represent unstable matrices.

Moving to $n=3$, our invariants now look as follows: $D = s_3, \delta
= 18 s_1 s_2 s_2 + s_1^2 s_2^2 - 4 s_2^3 - 4 s_1^3 s_3 - 27 s_3,
\Delta = s_1 s_2 - s_3$.  However, now we can no longer simply use their
signs to distinguish regions.  For instance, consider the regions
$R_{3000}$ and $R_{1200}$.  These both have $D > 0$ and $\delta > 0$
but to tell them apart, we need to note that the region $\delta > 0$
has two connected components and use the auxiliary condition that $s_1 > 0$ to
specify which of the components contains our region. The regions analogous to \ref{fig:region2x2} for three component systems where $D=1$ is given in \ref{fig:region3x3}.
%(Looking at the sign of $\Delta$ will not help here because, while $\Delta > 0$ in the region $R_{3000}$, it is also positive for some elements of $R_{1200}$ as well.)

\section{System graph and decomposition into strongly connected components}
Next, we will want to relate prpoperties of our matrix and its eigenvalues
to  properties of the associated graph $G_A$.  To do this, we will first
examine the category of paths on $G_A$, and derive a formula expressing
the chararacteristic polynomial as a sum over cycles.

Given a digraph $G$, let $\Path (a,b)$ denote the set of paths from the
vertex $a$ to the vertex $b$.  Given a path $p_1$ from $a$ to $b$ and a path
$p_2$ from $b$ to $c$, we may concatenate them to obtain a path $p_1 \circ
p_2$ from $a$ to $c$.  Hence, $(\Vertex(G), \Path(G), \circ)$ forms a category.

The characteristic polynomial $\chi(A)$ associated to $A$ can be factored
according to the strongly connected components of the graph $G$.
\[
\frac{\chi(M)'(z)}{\chi(M)(z)} = \tr (zI - M)^{-1}
\]
where, $\chi(M)(z) = \det (zI - M)$ is the chararacteristic polynomial of the
matrix $M$ with $z$ a complex variable.  This formula is readily verified when
$M$ is diagonal and has the nice property of being additive over decompositions
of the matrix. Making the change of variables $x=\frac{1}{z}$, we have
\[
\tr (I - xM)^n = \sum_{n=0}^\infty x^n \sum_{a \in \Vertex(G)} \sum_{p \in \Path_n (G)(a,a)} F(p)
\]
where $G$ is the graph of the matrix and $\Path_n (G;a,b)$ is the set of
paths of length $n$ from $a$ to $b$.  ; $F$ is the functor from this category
to the multiplicative semigroup of the real numbers which corresponds to the matrix $M$.

@@

>>>>

In order to accommodate the possibility of uncertainty in our modelling,
we will generalize our notion of a system on a network to that of a
system with random parameters.  Again, this will involve three
steps. First, we provide a measure space $S, \mathcal{B}(S)$ which represents the
values over which our parameters can vary. Next, instead of a single
vector field, we consider a family of vector fields parameterized by
this space. That is to say, we introduce a mapping $V \colon S \to \Gamma T(\prod_i M_m)$.
(Here $\Gamma$ denotes the space of sections of the tangent bundle.)
As before, for each point $p \in S$, the vector field $V(p)$ must be consistent with
the system graph. Finally, we select a probability measure $\mu \colon
\mathcal{B}(S) \to [0,1]$ which represents our understanding of which values of the parameters
are likely. In accord with Bayesian statistics, we might revise this
distribution as data comes in or use it to estimate parameters and error
bars from experimental results.

Having done this, we are now in a position to do a probabilistic analysis
of our system and its properties.  Given some quantity $q$ characterizing 
our system, this quantity becomes a random variable which may be discrete or 
continuous depending on the quantity under consideration.  For example, if 
the quantity is the time it takes for a particle to travel between two points, 
we have a real-valued random variable; if the quantity is the number of fixed
points, we have an integer-valued random variable; and if the quantity is
whether or not the system posesses a limit cycle, we have a binary random
variable.

@@

@@

\begin{proof}
Let $S_n$ denote the symmetric group on $n$ objects and let $f \colon
\mathbb{C}^n/S_n \to \mathbb{C}^n$ be the map that sends a set of roots
to the n-tuplet of coefficients of a polynomial.  This map is explicitly
given by the elemetary symmetric polynomials.  By the fundamental theorem
of algebra, it is invertible and, by ???, its inverse is continuous.

Suppose that $p$ is a stable polynomial.  By definition, this means that
$\Re r > 0$ for every root $r$ of $p$.  Since the set $\{r \mid \Re r > 0\}$
is an open subset of $\mathbb{C}^n/S_n$, the continuity of $f^{-1}$ implies 
that the set of stable polynomials is open.  Likewise, the set of unstable
polynomials none of whose roots have real part equal to zero is open.

Hence, if a point lies on the boundary of the stable region, it must 
correspond to a polynomial with at least one root whose real part is zero.
There are two ways this could happen.  If the root is real, then having
zero real part is the same as the root being zero, hence $s_n = 0$.  If
the root is complex, then having zero real part means that the root and
its complex conjugate are purely imaginary and sum to zero, hence $\Delta = 0$.
\end{proof}

\section{Paths and eigenvalues}
Next, we will want to relate prpoperties of our matrix and its eigenvalues
to  properties of the associated graph $G_A$.  To do this, we will first
examine the category of paths on $G_A$, and derive a formula expressing
the chararacteristic polynomial as a sum over cycles.

Given a digraph $G$, let $\Path (a,b)$ denote the set of paths from the
vertex $a$ to the vertex $b$ and let $\Path(G) = \bigcup_{a,b \in G} \Path (a,b)$. 
Given a path $p_1$ from $a$ to $b$ and a path $p_2$ from $b$ to $c$, we may 
concatenate them to obtain a path $p_1 \circ p_2$ from $a$ to $c$.  Since 
this operation is associative and has identity elements, 
$(\Vertex(G), \Path(G), \circ)$ forms a category.  Define 
$\mathcal{A}(G)$ to be the set of all linear combinations of elements of
$\Path(G)$ with coefficients in $\mathbb{R}$.  Then $\mathcal{A}(G)$ forms
an associative algebra under the multiplication induced by $\circ$.

% \def\wprod{\mathrm{wp}}
Next, we introduce two basic functors from $\Path (G)$.  One of these is
$\length \colon \Path(G) \to \mathbb{N}$ (regarding the natural numbers
as a category with a single object) which sends each path to its length.
The other functor is the weight product functor.  Given an assignement 
$w \colon \Path(G) \to \mathbb(R)$ of weights to the edges of our digraph,
define $\wprod \colon \Path(G) \to \mathbb{R}$ to be the function which sends
each path to the product of the weights associated to the edges which
that path traverses.  Given the way that graphs are associated to matrices,
weights on edges correspond to matrix entries and, given a path 
$v_1, v_2, v_3 \ldots$, we have $\wprod(v_1, v_2, v_3 \ldots) =
m_{v_1 v_2} m_{v_2 v_3} \cdots$.  Going in the opposite direction, we
note that any functor from $\Path(G)$ to $\mathbb{R}$ may be regarded as
a weighted product for a suitable choice of weights.  We extend $\wp$ to
$\mathcal{A}(G)$ by linearity; the result is an algebra homomorphism.

\todo{Start talking about Abelianizing the category and the algebra, say
more after decomposing into components.}

\todo{Characterize strongly connected components in terms of 
fibration over poset.}

The characteristic polynomial $\chi(A)$ associated to $A$ can be factored
according to the strongly connected components of the graph $G$.
\[
\frac{\chi(M)'(z)}{\chi(M)(z)} = \tr (zI - M)^{-1}
\]
where, $\chi(M)(z) = \det (zI - M)$ is the chararacteristic polynomial of the
matrix $M$ with $z$ a complex variable.  This formula is readily verified when
$M$ is diagonal and has the nice property of being additive over decompositions
of the matrix. Making the change of variables $x=\frac{1}{z}$, we have
\[
\tr (I - xM)^n =
\sum_{a \in \Vertex(G)} \sum_{p \in \Path_n (G)(a,a)} \wprod(p) x^{\length(p)}
\]
where $G$ is the graph of the matrix and $\Path_n (G;a,b)$ is the set of
paths of length $n$ from $a$ to $b$.  ; $F$ is the functor from this category
to the multiplicative semigroup of the real numbers which corresponds to the matrix $M$.

@@

\todo{Say a few words about Monte Carlo integration and error bars.}
